Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-10-25 manuscript/manuscript.tex
Modifications on manuscript

The main issue that should be overcome is taking into account the curvature of the Earth, thus the forward model must be defined in spherical coordinates.
A common way to achieve this is to discretize the Earth in spherical prisms known as tesseroids, which are defined by pairs of latitude, longitude and radius boundaries (see Figure \ref{fig:tesseroid-uieda}).
Calculating the gravity fields generated by an arbitrary tesseroid on any external point involves the resolution of volume integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion \citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

@@

In our case, we will assume that the tesseroid has an arbitrary variable density in depth, i.e. only depends on the radius spherical coordinate.
Thus, the integrals for the gravity fields are slightly modified:

@@

In the same way as for an homogeneous density, we can observe from equation \ref{eq:glq-var-dens} that the GLQ approximates the gravity fields of a tesseroid as the ones generated by point masses located at the nodes of the Legendre polynomials.
Nevertheless, in the variable density case, the information of the density function is summarised into the weights of the sum by taking into account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with well fitted discretization algorithms, approximates the gravity fields with good accuracy.\todo{no se si poner esta ultima oracion}

@@

Lets consider a spherical shell with inner and outer radii $R_1$ and $R_2$, respectivetly, whose density depends on the radius spherical coordinate (Figure \ref{fig:spherical-shell}).
The gravity potential generated by the shell on an arbitrary external point $Q(0,0,r)$, located along the $z$ axe at a distance $r$ from the origin, can be written as follows:

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/spherical-shell.pdf}
\caption{
Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
The computation point $Q$ is located in the $z$ axe at a distance $r$ from the centre of the shell.
For practical purposes we will assume that $Q$ is outside the outer radii, i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

@@

\noindent while the integration in $\phi'$ can be performed independently of the density function.
Taking into account the following integral:

\begin{equation}
    \int \frac{\cos \phi}{\sqrt{a - b \sin \phi}} \, d\phi =
    -\frac{2}{b} \sqrt{a - b \sin \phi} + C,
\end{equation}

\noindent the potential can be written as follows:

@@

For our purposes we can assume that the computation point $Q$ is outside the outer radii, i.e. $r>R_2$ ($R_2 \geq r'$), and simplify the square roots in equation \ref{eq:shell-pot-sqrts}:

@@

\noindent what leads to the following expression of the potential of the spherical shell:

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity potential generated by a spherical shell with a variable density in depth on any outside point located in the $z$ axe.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the integral in equation \ref{eq:shell-pot} reproduces the gravity potential field on any outside point at distance $r$ from the its centre.

@@

For practical purposes we are going to obtain the gravity potential for three different functions from the perspective of their variation in $r'$: a linear and an exponential density function.

@@

The second term on equation \ref{eq:shell-pot-linear} reproduces the potential generated by a spherical shell with homogeneous density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
So it can be read as the combination of a spherical shell with the later homogeneous density and the same shell with variable density $\rho(r') = ar'$.

@@

\noindent the potential can be written as follows:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
        Ab \, e^\frac{\Delta h}{b}
        \Big[
        & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1}{b}} - \\
        & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2}{b}}
        \Big].
    \end{split}
\label{eq:shell-pot-exp}
\end{equation}
}{
\begin{equation}
    V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
    Ab \, e^\frac{\Delta h}{b}
    \Big[
    & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1}{b}} - \\
    & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2}{b}}
    \Big].
\label{eq:shell-pot-exp}
\end{equation}
}

@@

This new library is based on the preexisting code developed by \citet{Uieda2016} (and included in Fatiando a Terra v.0.5, \citet{Uieda2013}) that is able to compute only the gravitational effects of tesseroids with homogeneous density.
\citet{Uieda2016} have optimised the more time consuming functions through the Numba compiler.
In contrast, we have decided to develop our new library in Cython language due to difficulties when passing a Python function as argument to precompiled Numba methods, keeping the homogeneous tesseroid calculation and adding the new variable density computations.
The Cython language is a superset of Python language that allows to generate efficient C code through the Cython compiler, producing a library whose functions can be called from regular Python code.

Moreover, this new software is compatible with the preexisting open-source library Fatiando a Terra and its usage is very similar to the one for homogeneous tesseroids:

@@

In a more detailed way, the new library approximates the gravitational effect of each tessseroid through the GLQ, and increases its accuracy with the modified version of the adaptive discretization algorithm, original developed by \citet{Li2011} and improved by \citet{Uieda2016}.

In addition, we had to develop a new density-based discretization algorithm in order to decrease a new type of error in the numerical approximation caused by high variations in the density functions.
It is intended to work together with the preexisting modified adaptive discretization algorithm.

@@

From equation \ref{eq:glq-var-dens} we saw that the GLQ approximates the volume integrals as the gravitational effect of $N_\lambda N_\phi N_r$ point masses, where $N_i$ are the orders of the quadrature for the integral on each coordinate, with $i \in \{ \lambda, \phi, r \}$.
If we set the GLQ orders to fixed values, it's easy to conclude that the nearer the computation point is to the tesseroid, the accuracy of the approximation is reduced and the point masses effect becomes more noticeable than the tesseroid we want to approximate.
A more complete analysis of this conclusion can be found in \citet{Ku1977, Li2011, Uieda2016}.

@@

If on the contrary the inequation \ref{eq:delta-density} is not held, then the tesseroid will not be subdivided.
The same happens if the density function is constant or if the tesseroid radius dimension is bellow the threshold of 1mm.
On the first case due to the fact that the density-based discretization algorithm is not needed, and on the second one because smaller tesseroids will not introduce meaningful increases in the accuracy of the computation.

@@

\noindent where $\rho_A$ and $\rho_B$ are constant densities, $R_1$ and $R_2$ are the inner and outer radii of the shell, respectively, and $R_c$ is the the radius on which the density change takes place.

@@

In the thin shell case we can conclude that a delta equal to 0.3 guarantee an acceptable accuracy under 0.1\% for the entire set of the density functions, while the thick shell case needs at least a delta equal to 0.2.
We will keep the more conservative value of 0.2 as the default value of delta on future computations.

>>>>

The main issue that should be overcome is taking into account the curvature of the Earth, thus the forward model must be defined in spherical coordinates.
A common way to achieve this is to discretize the Earth in spherical prisms known as tesseroids, which are defined by pairs of latitude, longitude and radial boundaries (see Figure \ref{fig:tesseroid-uieda}).
Calculating the gravity fields generated by an arbitrary tesseroid on any external point involves the resolution of volume integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion \citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

@@

In our case, we will assume that the tesseroid has an arbitrary variable density in depth, i.e. only depends on the radial spherical coordinate.
Thus, the integrals for the gravity fields are slightly modified:

@@

In the same way as for an homogeneous density, we can observe from equation \ref{eq:glq-var-dens} that the GLQ approximates the gravity fields of a tesseroid as the ones generated by $N_\lambda N_\phi N_r$ point masses located at the nodes of the Legendre polynomials, where $N_i$ are the orders of the quadrature for each integration, with $i \in \{ \lambda, \phi, r \}$.
Nevertheless, in the variable density case, the information of the density function is summarised into the weights of the sum by taking into account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with well fitted discretization algorithms, approximates the gravity fields with good accuracy.\todo{no se si poner esta ultima oracion}

@@

Lets consider a spherical shell with inner and outer radii $R_1$ and $R_2$, respectivetly, whose density depends on the radial spherical coordinate (Figure \ref{fig:spherical-shell}).
The gravity potential generated by the shell on an arbitrary external point $Q(0,0,r)$, located along the $z$ axe at a distance $r$ from the origin, can be written as follows:

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/spherical-shell.pdf}
\caption{
Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
The computation point $Q$ is located on the $z$ axe at a distance $r$ from the centre of the shell.
For practical purposes we will assume that $Q$ is outside the outer radii, i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

@@

\noindent while the integration in $\phi'$ can be performed independently of the density function.
Making use of SymPy \citep{sympy2017}, a Python library for symbolic mathematics, we obtained the following expression for the potential:

@@

For our purposes we can assume that the computation point $Q$ is farther from the origin than the outer radius, i.e. $r>R_2$. 
And taking into account that $r'$ is always lower than $R_2$, we can simplify the square roots in equation \ref{eq:shell-pot-sqrts}:

@@

\noindent what leads to the following expression of the potential:

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact gravity potential generated by a spherical shell with a variable density in depth on any outside point located on the $z$ axe.
Due to the rotational symmetry along any axe that passes through the centre of the shell, the integral in equation \ref{eq:shell-pot} reproduces the gravity potential field on any outside point at distance $r$ from the its centre.

@@

For future purposes we are going to obtain the gravity potential for three different functions from the perspective of their variation in $r'$: a linear and an exponential density function.

@@

The first term on equation \ref{eq:shell-pot-linear} reproduces the potential generated by a spherical shell with variable density $\rho(r') = ar'$, while the second one constitutes the potential generated by a spherical shell with homogeneous density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
So the potential can be read as the one generated by a combination of both shells.

@@

\noindent the potential (calculated through SymPy \citep{sympy2017}) can be written as follows:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
        Ab 
        \Big[
        & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1 - \Delta h}{b}} - \\
        & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2 - \Delta h}{b}}
        \Big].
    \end{split}
\label{eq:shell-pot-exp}
\end{equation}
}{
\begin{equation}
    V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
    Ab \, e^\frac{\Delta h}{b}
    \Big[
    & (R_1^2 + 2R_1 b + 2b^2)e^{-\frac{R_1}{b}} - \\
    & (R_2^2 + 2R_2 b + 2b^2)e^{-\frac{R_2}{b}}
    \Big].
\label{eq:shell-pot-exp}
\end{equation}
}

@@

This new library approximates the gravitational effect of each tesseroid through the GLQ, and guarantees its accuracy with the modified version of the adaptive discretization algorithm, original developed by \citet{Li2011} and improved by \citet{Uieda2016}, and a new density-based discretization algorithm.
The later one had to be developed in order to decrease a new type of error in the numerical approximation caused by high variations in the density functions.

Moreover, the new forward model is based on the preexisting code developed by \citet{Uieda2016} (and included in Fatiando a Terra v.0.5, \citet{Uieda2013}) that is able to compute the gravitational effects of tesseroids with homogeneous density.
On that implementation \citet{Uieda2016} have optimised the more time consuming functions through the Numba compiler.
In contrast, we have decided to develop our new library in Cython language due to difficulties when passing a Python function as argument to precompiled Numba methods, keeping the homogeneous tesseroid calculation and adding the new variable density computations.
The Cython language is a superset of Python language that allows to generate efficient C code through the Cython compiler, producing a library whose functions can be called from regular Python code.

It is also compatible with the preexisting open-source library Fatiando a Terra and its usage is very similar to the one for homogeneous tesseroids:

@@

@@

From equation \ref{eq:glq-var-dens} we saw that the GLQ approximates the volume integrals as the gravitational effect of point masses located at the nodes of the quadrature.
If we set the GLQ orders to fixed values, it's easy to conclude that the nearer the computation point is to the tesseroid, the accuracy of the approximation is reduced and the point masses effect becomes more noticeable than the tesseroid we want to approximate.
A more complete analysis of this conclusion can be found in \citet{Ku1977, Li2011, Uieda2016}.

@@

If on the contrary the inequation \ref{eq:delta-density} is not held, then the tesseroid will not be subdivided.
The same happens if the density function is constant or if the tesseroid radial dimension is bellow the threshold of 1mm.
On the first case due to the fact that the density-based discretization algorithm is not needed, and on the second one because smaller tesseroids will not introduce meaningful increases in the accuracy of the computation.

@@

\noindent where $\rho_A$ and $\rho_B$ are constant densities, $R_1$ and $R_2$ are the inner and outer radii of the shell, respectively, and $R_c$ is the radius on which the density change takes place.

@@

In the thin shell case we can conclude that a delta equal to 0.3 guarantees an acceptable accuracy under 0.1\% for the entire set of the density functions, while the thick shell case needs at least a delta equal to 0.2.
We will keep the more conservative value of 0.2 as the default value of delta on future computations.
